Solving \\ the \\ 15-Puzzle \\ \& \\ Rubik's Cube

### 2022-12-17 latex/Chapters/Puzzles.tex
typos

\Subsection{History -- from Erno Rubik to speed cubing competitions}
The second puzzle I will focus on is the Rubik's Cube (\textbf{RC}), invented in 1974 by Erno Rubik, originally known as \textit{Magic Cube} (see \cite{RubiksWiki}). Since it was commercialised in 1980, it has known a worldwide success: countless people are playing with the original 3x3x3 cube, as well as with variants of it of all shapes and sizes. Competitions bring together \textit{speedcubers} who have trained for years and memorized algorithms to solve the Rubik's in astonishingly fast times (seconds for the best ones). Some of the principles used by speedcubers generalise to different dimensions. As an example, it is quite impressive to watch \textit{Cubastic} solve a 15x15x15 cube in (only) two and a half hours! (see \cite{151515Rubiks}). The key insight to solving high dimensional \textbf{RC}s (short obviously of coming up with specific strategies) is the following: if we manage to group all the inside (i.e. not located on a corner or an edge) cubies by colour, and then all the non-corner edges on a given tranche of a face by colour, we are essentially reducing the cube to a 3x3x3, since no face rotation will ever undo these groupings. The picture below (\ref{fig:4442333}) shows my own 4x4x4 and 3x3x3 cubes arranged to be in identical configurations.
\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.25]{./Figures/RC333444}}
  \caption[Reducing RC]{Reducing a 4x4x4 \textbf{RC} to a 3x3x3 -- home-made picture}
  \label{fig:4442333}
\end{figure}
\noindent If we can bring the 4x4x4 cube to such a configuration on all faces (which is not very hard), and if we know how to solve a 3x3x3, we are essentially done. (The additional degrees of freedom on the 4x4x4 complicate slightly the story since certain arrangements of colors which are possible on the 4x4x4 are not on the 3x3x3, e.g. red an and white on opposite faces). Overall, the difficulty of hand-solving cubes does not grow as much with dimensionality as one would think, and the time difference comes mostly from the difficulty of handling and manipulating larger cubes.

\Subsection{Conventions: faces, moves and configurations}
I follow the conventions and notations of the \textbf{RC} community and refer to the faces as F (front), B (back), U (up), D (down), L (left) and R (right). It is common in \textbf{RC} jargon to refer to moves  by the name of the face that is rotated clockwise (when facing it), and by adding a \textit{'} for counter-clockwise rotations. Sometimes numbers are used to indicate repeated moves, though I will count that as multiple moves, since obviously e.g. U2 = U U = U' U'. As another example: F B' F2 means rotating the front face clockwise, the back face counter-clockwise, followed by the front face twice. The convention I follow is called the \textit{quarter-turn-metric}, and obviously the length of solutions is affected by which metric we use.
\\
Since I am interested in experimenting with \textit{no-human-knowledge} methods such as \textbf{DRL} and \textbf{DQL}, I will consider cubes which are identical up to full cube rotation such as the \textit{flattened} cubes on figure \ref{fig:configurations} ) to be different. They will be represented by different tensors in my code, and this is up to the solvers to make sense of things.

>>>>

\Subsection{History -- from Erno Rubik to speed cubing competitions}
The second puzzle I will focus on is the Rubik's Cube (\textbf{RC}), invented in 1974 by Erno Rubik, originally known as \textit{Magic Cube} (see \cite{RubiksWiki}). Since it was commercialised in 1980, it has known a worldwide success: countless people are playing with the original 3x3x3 cube, as well as with variants of it of all shapes and sizes. Competitions bring together \textit{speedcubers} who have trained for years and memorized algorithms to solve the Rubik's in astonishingly fast times (seconds for the best ones). Some of the principles used by speedcubers generalise to different dimensions. As an example, it is quite impressive to watch \textit{Cubastic} solve a 15x15x15 cube in (only) two and a half hours! (see \cite{151515Rubiks}). The key insight to solving high dimensional \textbf{RC}s (short obviously of coming up with specific strategies) is the following: if we manage to group all the inside (i.e. not located on a corner or an edge) cubies by colour, and then all the non-corner edges on a given tranche of a face by colour, we are essentially reducing the cube to a 3x3x3, since no face rotation will ever undo these groupings. The picture below (\ref{fig:4442333}) shows my own 4x4x4 and 3x3x3 cubes arranged to be in identical configurations.
\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.25]{./Figures/RC333444}}
  \caption[Reducing RC]{Reducing a 4x4x4 \textbf{RC} to a 3x3x3 -- home-made picture}
  \label{fig:4442333}
\end{figure}
\noindent If we can bring the 4x4x4 cube to such a configuration on all faces (which is not very hard), and if we know how to solve a 3x3x3, we are essentially done. (The additional degrees of freedom on the 4x4x4 complicate slightly the story since certain arrangements of colors which are possible on the 4x4x4 are not on the 3x3x3, e.g. red and white on opposite faces). Overall, the difficulty of hand-solving cubes does not grow as much with dimensionality as one would think, and the time difference comes mostly from the difficulty of handling and manipulating larger cubes.

\Subsection{Conventions: faces, moves and configurations}
I follow the conventions and notations of the \textbf{RC} community and refer to the faces as F (front), B (back), U (up), D (down), L (left) and R (right). It is common in \textbf{RC} jargon to refer to moves  by the name of the face that is rotated clockwise (when facing it), and by adding a \textit{'} for counter-clockwise rotations. Sometimes numbers are used to indicate repeated moves, though I will count that as multiple moves, since obviously e.g. U2 = U U = U' U'. As another example: F B' F2 means rotating the front face clockwise, the back face counter-clockwise, followed by the front face twice. The convention I follow is called the \textit{quarter-turn-metric}, and obviously the length of solutions is affected by which metric we use.
\\
Since I am interested in experimenting with \textit{no-human-knowledge} methods such as \textbf{DRL} and \textbf{DQL}, I will consider cubes which are identical up to full cube rotation such as the \textit{flattened} cubes on figure \ref{fig:configurations} to be different. They will be represented by different tensors in my code, and this is up to the solvers to make sense of things.

### 2022-12-17 latex/Chapters/Rubiks.tex
last cleanup
typos

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.4]{./Figures//222RCDRL}}
  \caption[222RCDRL]{\textbf{DRL} of the 2x2x2 \textbf{RC}}
  \label{fig:222RCDRL}
\end{figure}

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.4]{./Figures//222RCDQL}}
  \caption[222RCDQL]{\textbf{DQL} of the 2x2x2 \textbf{RC}}
  \label{fig:222RCDQL}
\end{figure}

@@

I summarize as usual the 2x2x2 \textbf{RC} performance tests with my usual metrics in figure \ref{fig:222RCPerformance}. Overall, the \textbf{DL} and \textbf{DRL} managed to solve all cubes (though \textbf{DL} took over an hour and \textbf{DRL} 6 hours for their slowest cube. Both methods outperformed Kociemba in terms of optimality-score and got on par with A$^{*}[Kociemba]$. Their max cost over all cubes encountered was 12 for up to 20 shuffles, and 13 for $\infty$ shuffling. Considering God's number for the 2x2x2 \textbf{RC} of 11, this is rather impressive!
\\
\\
Quite remarkably we can see (second graph of \ref{fig:222RCDRL}) that the \textbf{DRL} heuristic only ever assigned a cost-to-go of 7.6 at most. Despite that, it was still able to solve cubes requiring up to 13 moves (likely to have been among the hardest 2x2x2 cubes). How is that possible? Notice that the loss of 0.2 towards which the network converged at the end of its training is misleading, since this is in-sample loss versus the target network. My theory is that while the value the \textbf{DRL} assigned to hard-to-solve cubes can be off (by at least 4 or 5), it is likely that the network gets the relative value of neighbouring cubes right (not that surprising given the value-iteration), and this is sufficient to guide A$^{*}$ extremely well. If both the absolute value of the cost-to-go and the relative (among neighbouring cubes) was totally off, we would indeed not expect the heuristic to make A$^{*}$ any better than, say, \textbf{BFS} and we would therefore not expect it to be able to solve \textit{costly} cubes requiring 10+ moves.
\\
\\
Finally, we can see that \textbf{DQL}, despite showing visually cleaner and more monotonic loss convergence than \textbf{DRL} (see \ref{fig:222RCDQL} versus \ref{fig:222RCDRL}), struggled a lot more to solve difficult cubes (above 10 shuffles). This shows that my earlier findings on the 3x3 \textbf{SP} are likely not universal, and very much depend on the parameters set during training, as well as on the problem at hand.

>>>>

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.4]{./Figures/222RCDRL}}
  \caption[222RCDRL]{\textbf{DRL} of the 2x2x2 \textbf{RC}}
  \label{fig:222RCDRL}
\end{figure}

\begin{figure}[H]
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.4]{./Figures/222RCDQL}}
  \caption[222RCDQL]{\textbf{DQL} of the 2x2x2 \textbf{RC}}
  \label{fig:222RCDQL}
\end{figure}

@@

I summarize as usual the 2x2x2 \textbf{RC} performance tests with my usual metrics in figure \ref{fig:222RCPerformance}. Overall, the \textbf{DL} and \textbf{DRL} managed to solve all cubes (though \textbf{DL} took over an hour and \textbf{DRL} 6 hours for their slowest cube. Both methods outperformed Kociemba in terms of optimality-score and got on par with A$^{*}[Kociemba]$. Their max cost over all cubes encountered was 12 for up to 20 shuffles, and 13 for $\infty$ shuffling. Considering God's number for the 2x2x2 \textbf{RC} of 11, this is rather impressive!
\\
\\
Quite remarkably we can see (second graph of \ref{fig:222RCDRL}) that the \textbf{DRL} heuristic only ever assigned a cost-to-go of 7.6 at most. Despite that, it was still able to solve cubes requiring up to 13 moves (likely to have been among the hardest 2x2x2 cubes). How is that possible? Notice that the loss of 0.2 towards which the network converged at the end of its training is misleading, since this is in-sample loss versus the target network. My theory is that while the value the \textbf{DRL} network assigned to hard-to-solve cubes can be off (by at least 4 or 5), it is likely that it gets the relative value of neighbouring cubes right (not that surprising given value-iteration assigns values from that of neighbours!), and this is sufficient to guide A$^{*}$ extremely well. If both the absolute value of the cost-to-go and the relative (among neighbouring cubes) was totally off, we would indeed not expect the heuristic to make A$^{*}$ any better than, say, \textbf{BFS} and we would therefore not expect it to be able to solve \textit{costly} cubes requiring 10+ moves.
\\
\\
Finally, we can see that \textbf{DQL}, despite showing visually cleaner and more monotonic loss convergence than \textbf{DRL} (see \ref{fig:222RCDQL} versus \ref{fig:222RCDRL}), struggled a lot more to solve difficult cubes (above 10 shuffles). This shows that my earlier findings on the 3x3 \textbf{SP} are likely not universal, and very much depend on the parameters set during training, as well as on the problem at hand.

### 2022-12-17 latex/Chapters/SlidingPuzzle.tex
typos

\noindent I am not making claims of depth (each solver could be tuned and optimized much further) nor breadth (I have not run all solvers versus all dimensions). These experiments take a lot of computing time, so I had to be somewhat selective. I wanted however to run all of the solvers on at least one particular dimension (3x3 seemed appropriate) to try and answer, even if not in generality, as many of the following interesting questions:
\begin{itemize}
\item How hard is too hard for \textbf{BFS} to complete in \textit{reasonable} time?
\item How much, if any, improvement do we get with Manhattan++ over Manhattan?
\item Does \textbf{DL} perform much better than \textbf{DRL} due to supervision?
\item Is \textbf{DQL}, all things being equal, able to learn a better value-heuristic than \textbf{DRL}?
\item How does \textbf{MCTS} perform? How important is it to tune the hyper-parameter $c$? How much improvement does the final \textbf{BFS} tree-trimming bring?
\item Are the \textbf{D*L} heuristics competitive versus Manhattan++ or the perfect heuristic?
\end{itemize}

@@

\Subsection{Solvers' comparison}
\label{ssec:33SPSC}
\label{sec:S33DRLDQL}
As promised earlier in table \ref{tab:gridSP}, I have run the Solver's comparison test with 10 different solvers (Naive, \textbf{MCTS}, \textbf{BFS}, A$^{*}$ with the following heuristics: Manhattan, Manhattan++, Perfect, \textbf{DRL}, \text{DQL} and \text{DL}), to get an idea of their respective success rate, strengths and weaknesses. The \textbf{DRL} and \textbf{DQL} heuristics were trained with the exact same parameters and a fully connected network of 3 hidden layers with 600, 300 and 100 nodes. For the \textbf{DL} heuristic (trained on perfect solutions from A$^{*}$[Perfect]) I tried with the same fully connected network as well as with a combination of parallel fully connected and convolutional layers (see figure \ref{fig:33SPnets} for the exact architecture I used). The detailed results (percentage of configurations solved, optimality score, median and max cost, run time and expanded nodes) can be found on the next page  in figure \ref{fig:33SPPerformance}. Each solver had to solve 1,000 random puzzles for each level of difficulty (scrambling from goal) going from 5 to 50 in step of 5, as well as for perfect shuffling. Let me list some of the important findings from this rather comprehensive experiment:

\begin{itemize}
\item As expected, \textbf{BFS} becomes quickly hopeless.
\item The naive solver is obviously by far the fastest but finds poor solutions (about 2.5 longer on average than optimal).
\item As discussed in the previous section, \textbf{MCTS} is slow, and even with much exploration and \textbf{BFS} tree trimming found solution about 25\% longer than optimal.
\item I originally was naively hoping a pure convolutional network would do well but did not manage to get the loss to converge. I later realized that this is quite different from the typical image patterns recognition as there is less direct translation and scale invariants at play here, and decided instead to try combining convolutional with fully connected layers in parallel in order to appropriately value 2x2 patterns together with their position on the sliding-puzzle. That did improve things, but overall I saw little difference in performance from my simpler 3-hidden-layer fully connected network.
\item \textbf{DRL} and \textbf{DQL} heuristics, though not as efficient in terms node-expansion as the perfect and the \textbf{DL} heuristics, were rather surprisingly doing better than Manhattan++. They ended up \textit{seeing} around 25\% of all the possible configurations during training (in an \textit{unsupervised} way though) so did show some reasonable amount of out-of-sample generalization. Their optimality score was an impressive 99\%+ at all levels of difficulty for \textbf{DRL} and 98\%+ for \textbf{DQL}. Even more interestingly, the \textbf{DQL} heuristic expanded almost 4 times less nodes than \textbf{DRL}, suggesting some benefit from the joint learning of the policy and value function.
\item The \textbf{DL} heuristics (fully connected and convolutional) performed almost on par with the perfect heuristic (optimality score of 99\%+ at all levels of difficulty and nodes expansion only increased by about 12\%), despite seeing only 8\% of the configurations by the end of their training.
\end{itemize}

>>>>

\noindent I am not making claims of depth (each solver could be tuned and optimized much further) nor breadth (I have not run all solvers versus all dimensions). These experiments take a lot of computing time, so I had to be somewhat selective. I wanted however to run all of the solvers on at least one particular dimension (3x3 seemed appropriate) to try and answer, even if not in generality, as many of the following interesting questions:
\begin{itemize}
\item How hard is too hard for \textbf{BFS} to complete in \textit{reasonable} time?
\item How much, if any, improvement do we get with Manhattan++ over Manhattan?
\item Does \textbf{DL} perform much better than \textbf{DRL} due to supervision?
\item Is \textbf{DQL}, all things being equal, able to learn a better value-heuristic than \textbf{DRL}?
\item How does \textbf{MCTS} perform? How important is it to tune the hyper-parameter $c$? How much improvement does the final \textbf{BFS} tree-trimming bring?
\item Are the \textbf{DxL} heuristics competitive versus Manhattan++ or the perfect heuristic?
\end{itemize}

@@

\Subsection{Solvers' comparison}
\label{ssec:33SPSC}
\label{sec:S33DRLDQL}
As promised earlier in table \ref{tab:gridSP}, I have run the Solver's comparison test with 10 different solvers (Naive, \textbf{MCTS}, \textbf{BFS}, A$^{*}$ with the following heuristics: Manhattan, Manhattan++, Perfect, \textbf{DRL}, \textbf{DQL} and \textbf{DL}), to get an idea of their respective success rate, strengths and weaknesses. The \textbf{DRL} and \textbf{DQL} heuristics were trained with the exact same parameters and a fully connected network of 3 hidden layers with 600, 300 and 100 nodes. For the \textbf{DL} heuristic (trained on perfect solutions from A$^{*}$[Perfect]) I tried with the same fully connected network as well as with a combination of parallel fully connected and convolutional layers (see figure \ref{fig:33SPnets} for the exact architecture I used). The detailed results (percentage of configurations solved, optimality score, median and max cost, run time and expanded nodes) can be found on the next page  in figure \ref{fig:33SPPerformance}. Each solver had to solve 1,000 random puzzles for each level of difficulty (scrambling from goal) going from 5 to 50 in step of 5, as well as for perfect shuffling. Let me list some of the important findings from this rather comprehensive experiment:

\begin{itemize}
\item As expected, \textbf{BFS} becomes quickly hopeless.
\item The naive solver is obviously by far the fastest but finds poor solutions (about 2.5 longer on average than optimal).
\item As discussed in the previous section, \textbf{MCTS} is slow, and even with much exploration and \textbf{BFS} tree trimming found solution about 25\% longer than optimal.
\item I originally was naively hoping a pure convolutional network would do well but did not manage to get the loss to converge. I later realized that this is quite different from the typical image patterns recognition as there is less direct translation and scale invariants at play here, and decided instead to try combining convolutional with fully connected layers in parallel in order to appropriately value 2x2 patterns together with their position on the sliding-puzzle. That did improve things, but overall I saw little difference in performance from my simpler 3-hidden-layer fully connected network.
\item \textbf{DRL} and \textbf{DQL} heuristics, though not as efficient in terms of node-expansion as the perfect and the \textbf{DL} heuristics, were rather surprisingly doing better than Manhattan++. They ended up \textit{seeing} around 25\% of all the possible configurations during training (in an \textit{unsupervised} way though) so did show some reasonable amount of out-of-sample generalization. Their optimality score was an impressive 99\%+ at all levels of difficulty for \textbf{DRL} and 98\%+ for \textbf{DQL}. Even more interestingly, the \textbf{DQL} heuristic expanded almost 4 times less nodes than \textbf{DRL}, suggesting some benefit from the joint learning of the policy and value function.
\item The \textbf{DL} heuristics (fully connected and convolutional) performed almost on par with the perfect heuristic (optimality score of 99\%+ at all levels of difficulty and nodes expansion only increased by about 12\%), despite seeing only 8\% of the configurations by the end of their training.
\end{itemize}
